---
title: "Most harmful and damging weather events in the US since 2000"
author: "Johan van Tent"
date: "2019-10-30"
output:
  html_document:
    code_folding: "show"

---

```{r setup, echo = TRUE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
suppressPackageStartupMessages(library(tidyverse))

# Questions
# 
# Your data analysis must address the following questions:
# 
# 1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most
#    harmful with respect to population health?
# 2. Across the United States, which types of events have the greatest economic consequences?

```

## Synopsis

This analysis explores the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. The first recorded weather events in the database go back to 1950 but most events are recorded in recent years. The analysis focuses on the type of weather events that have caused the most harm to humans and the most economic damage since 2000. Heat causes the most fatalities, but tornadoes are reported to cause the most injuries. Floods cause the largest amount of economic damage. Efforts to prevent damage or harm to humans should probably focus most on these types of weather events.


## Data Processing

### Downloading and reading

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

The Storm Dataset can be downloaded using this [link](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2).

```{r downloading-data}

# download data if necessary - but only once

if (!dir.exists("data")) dir.create("data")

if (!file.exists("data/storm_data.bz2")) { 
  download.file(url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
                destfile = "data/storm_data.bz2")
}
```

The Storm Dataset comes as compressed csv-file. It can be read into a dataframe using the `read_csv()`-function.

```{r reading-data, cache=TRUE}

# readr::read_csv can read zipped files directly

storm_data <- read_csv("data/storm_data.bz2",
                col_types = cols(BGN_DATE = col_date("%m/%d/%Y %H:%M:%S"),
                                 PROPDMGEXP = col_character(),
                                 CROPDMGEXP = col_character()))
# # temporary
 # storm_data <- read_csv("data/storm_data",
 #                  col_types = cols(BGN_DATE = col_date("%m/%d/%Y %H:%M:%S"))) 

```

### Inspecting the data

```{r inspecting-data}

n_rows <- nrow(storm_data)
n_vars <- ncol(storm_data)

n_event_types <- length(unique(storm_data$EVTYPE))

```

The storm dataset has:

- `r n_vars` variables
- `r n_rows` rows 
- `r n_event_types` unique event labels

The first events were recorded in 1950. In general more events are recorded in recent years. This is shown in the graph below.

```{r event-count, fig.cap="*Number of recorded events in the storm dataset*", fig.width = 8}
  storm_data %>% 
  mutate(year = lubridate::year(BGN_DATE)) %>% 
  ggplot(aes(year)) + 
  geom_bar() + 
  labs(title = "Number of events per year")
```



### Transformations

In the variable `EVTYPE` there seem to be a number of different labels that refer to the same type of event, e.g. *THUNDERSTORM WIND* and *THUNDERSTORM WINDS*. It is therefore necessary to clean the labels of the events. For this analysis we're only interested in events that cause a lot of injuries or damage. Therefore we'll only relabel those events that occur often or cause a lot of damage or injuries.

We're interested in preventing damage and casualties for future events. It seems reasonable to think that only recent events are relevant for this purpose. Therefore we choose to only include data from 2000 or later in the analysis. Apart from that, it seems also likely that a lot of events are missing in earlier years.

Damage is expressed in thousands (K), millions (M) or billions (B). To be able to compare the amount of damage it is necessary to convert the damage to the same unit.

```{r transformations}

relabel_df <- tibble::tribble(
               ~EVTYPE,         ~EVTYPE_NEW,
  "THUNDERSTORM WINDS", "THUNDERSTORM WIND",
           "TSTM WIND", "THUNDERSTORM WIND",
      "EXCESSIVE HEAT",              "HEAT"
  )

storm_data_trans <- 
  storm_data %>% 
  # relabelling
  left_join(relabel_df, by = "EVTYPE") %>% 
  mutate(EVTYPE = ifelse(is.na(EVTYPE_NEW), EVTYPE, EVTYPE_NEW)) %>% 
  
  # selecting only data from 2000 and later
  filter(lubridate::year(BGN_DATE) >= 2000) %>% 
  
  # scaling the amount of damges all to the same unit of single dollars ($)
  mutate(CROPDMG = case_when(
    CROPDMGEXP == "K" ~ CROPDMG * 1000,
    CROPDMGEXP == "M" ~ CROPDMG * 1000 * 1000,
    CROPDMGEXP == "B" ~ CROPDMG * 1000 * 1000 * 1000,
    TRUE ~ CROPDMG),
         PROPDMG = case_when(
    PROPDMGEXP == "K" ~ PROPDMG * 1000,
    PROPDMGEXP == "M" ~ PROPDMG * 1000 * 1000,
    PROPDMGEXP == "B" ~ PROPDMG * 1000 * 1000 * 1000,
    TRUE ~ PROPDMG)
    )

```


## Results

The storm database con

### Harmful events

The figure below shows the total amount of fatalities and injuries caused by weather events since 2000. Only the top-10 types of events that have caused the most fatalities are included.

```{r harmful-events, fig.width = 10, fig.cap="*The 10 most harmful events since 2000*"}
# Selecting only events with harmful effects for people: At least 1 fatality or 1 injury. 
harmful_events <- storm_data_trans %>% 
  filter(FATALITIES > 0 | INJURIES > 0)


# Determining and plotting the most harmful events
harmful_events %>% 
  group_by(EVTYPE) %>% 
  summarise(Fatalities = sum(FATALITIES, na.rm = TRUE), 
            Injuries = sum(INJURIES, na.rm = TRUE)) %>% 
  ungroup() %>% 
  arrange(desc(Fatalities)) %>% 
  top_n(10, wt = Fatalities) %>% 
  gather("harm_type", "n", -EVTYPE) %>% 
  ggplot(aes(fct_rev(fct_inorder(EVTYPE)), n)) + 
  geom_col() + 
  facet_wrap(~harm_type, scales = "free") +
  labs(title = "Most harmful event-types since 2000",
       x = "",
       y = "Number of casualties") +
  coord_flip()


```

It can be seen that *heat* caused the most fatalities although *tornadoes* have caused almost as much fatalities. Both heat and tornadoes cause about twice or more fatalities than other types of events. Tornadoes are also reponsible for a large amount of injuries: thrice or more the amount of injuries caused by other events like heat or thunderstorm wind.

Efforts to prevent harm to humans from weather events should probably focus most on heat and tornadoes.

### Economic damage

The figure below shows the total amount of damage in billions of dollars caused by weather events since 2000. Only the top-10 types of events that have caused the most damage are included.

```{r damaging-events, fig.width = 8, fig.cap="*The 10 most damaging types of events*"}

damaging_events <- storm_data_trans %>% 
  filter(PROPDMG > 0 | CROPDMG > 0) %>% 
  mutate(DAMAGE = PROPDMG + CROPDMG)

damaging_events %>% 
  group_by(EVTYPE) %>% 
  summarise(DAMAGE = sum(DAMAGE, na.rm = TRUE)) %>% 
  ungroup() %>% 
  arrange(desc(DAMAGE)) %>% 
  top_n(10, wt = DAMAGE) %>% 
  ggplot(aes(fct_rev(fct_inorder(EVTYPE)), DAMAGE / 1000000000)) + 
  geom_col() + 
  labs(title = "Most damaging event-types since 2000",
       x = "",
       y = "Billions of Dollars") +
  coord_flip()


```

From the figure it can be seen that floods cause by far the most damage, almost twice as much than the next type of event hurricanes. Hurricanes in turn cause more than twice as much damage than other weather events, floods excluded of course.

Efforts to prevent economic damage should probably focus most on floods and hurricanes.
